### 2015-04-18 motivation.tex
edits to motivation

To cope with software diversity, many HPC sites use a combination of existing
packages managers and either manual or semi-automated software installation.  
For the baseline OS, many sites maintain traditional binary
packages using the vendor's package manager. LLNL maintains a Linux 
distribution, CHAOS~\cite{chaos} for this purpose.

@@

Diverse software versions not only present problems for build and installation;
they also complicate the runtime environment. When launched, an executable
must determine the location of its dependency libraries, or it will not run.
Statically linked binaries do not have this issue, but modern
operating systems make extensive use of dynamic linking.
By default, the dynamic loader on most systems is configured to search only
system library paths such as {\tt /lib}, {\tt /usr/lib}, and
{\tt /usr/local/lib}.  If binaries are installed in other locations, the
{\it user} who runs the program (often not the user or staff member who
installed the binary) must typically add dependency library paths to
{\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
can find them.

@@

The alternative to per-user environment settings is to embed library search 
paths in installed binaries at compile time. When set this way, the search 
path is called an {\tt RPATH}. {\tt RPATHs} and environment modules are not 
mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
and variables, but adding {\tt RPATHs} ensures that binaries run correctly 
regardless of whether the right module is loaded. LC installs software with 
both {\tt RPATHs} and {\tt dotkit}~\cite{dotkit} modules.

@@

Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
Like Nix, it uses cryptographic versioning, and it archives installations
in a git repository so that they can be recalled on demand. 
%
Both Nix and hashdist use {\tt RPATHs} in their packages to ensure that
libraries are found correctly.

@@

\paragraph{Gaps in current common practice}
The cryptographic versioning of Nix and Hashdist is very flexible. It versions
the package {\it and} its dependency configuration, and can represent any
configuration. However, users cannot navigate or query the installed software.
The systems are in a sense ``write-only'':
Nix does not even offer a way to ask for a package's dependencies.
%
EasyBuild and Smithy generate environment modules, which allow querying
through their own interface. Naming in existing module systems, however, cannot
handle combinatoric versions. The Lmod authors call this the
``matrix problem``~\cite{mclay:lmod-tutorial}.

@@

The main limitation of existing tools is the lack of build {\it composability}.
The full set of package versions is combinatoric, and realizing an arbitrary
combination of compiler, MPI version, build options, and dependency versions
requires tediously modifying many package files.
%
Indeed, the number of package files required for existing systems scales with
the number of version {\it combinations}, not the number of packages, which.
quickly becomes unmanageable.  As an example, the very flexible EasyBuild
system has over 3,300 files for several permutations of only 600 packages. 
Other systems have similar issues. HPC sites need a mechanism to 
{\it parameterize} packages so that new builds can be {\it composed} on the fly.

>>>>

To cope with software diversity, many HPC sites use a combination of existing
package managers and either manual or semi-automated software installation.  
For the baseline OS, many sites maintain traditional binary
packages using the vendor's package manager. LLNL maintains a Linux 
distribution, CHAOS~\cite{chaos} for this purpose, which is managed by RPM.

@@

Diverse software versions not only present problems for build and installation;
they also complicate the runtime environment. When launched, an executable
must determine the location of its dependency libraries, or it will not run,
or even worse, it may find the wrong dependencies and subtly produce incorrect results.
Statically linked binaries do not have this issue, but modern
operating systems make extensive use of dynamic linking.
By default, the dynamic loader on most systems is configured to search only
system library paths such as {\tt /lib}, {\tt /usr/lib}, and
{\tt /usr/local/lib}.  If binaries are installed in other locations, the
{\it user} who runs the program (often not the same person who
installed the binary) must typically add dependency library paths to
{\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
can find them.

@@

The alternative to per-user environment settings is to embed library search 
paths in installed binaries at compile time. When set this way, the search 
path is called an {\tt RPATH}. {\tt RPATHs} and environment modules are not 
mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
and other variables, but adding {\tt RPATHs} ensures that binaries run correctly 
regardless of whether the right module is loaded. LC installs software with 
both {\tt RPATHs} and {\tt dotkit}~\cite{dotkit} modules.

@@

HashDist~\cite{hashdist} is a meta-build system and package manager for HPC.
Like Nix, it uses cryptographic versioning, and it archives installations
in a git repository so that they can be recalled on demand. 
%
Both Nix and HashDist use {\tt RPATHs} in their packages to ensure that
libraries are found correctly.

@@

\paragraph{Gaps in current common practice}
The cryptographic versioning of Nix and HashDist is very flexible. It versions
the package {\it and} its dependency configuration, and can represent any
configuration. However, users cannot navigate or query the installed software.
The systems are in a sense ``write-only'':
Nix does not even offer a way to ask for a package's dependencies.
%
EasyBuild and Smithy generate environment modules, which allow querying
through their own interface. Naming schemes used in existing module systems, however, cannot
handle combinatoric versions. The Lmod authors call this the
``matrix problem``~\cite{mclay:lmod-tutorial}.

@@

The main limitation of existing tools is the lack of build {\it composability}.
The full set of package versions is combinatoric, and realizing an arbitrary
combination of compiler, MPI version, build options, and dependency versions
requires tediously modifying many package files.
%
Indeed, the number of package files required for existing systems scales with
the number of version {\it combinations}, not the number of packages, which
quickly becomes unmanageable.  As an example, the very flexible EasyBuild
system has over 3,300 files for several permutations of only 600 packages. 
Other systems have similar issues. HPC sites need a mechanism to 
{\it parameterize} packages so that new builds can be {\it composed} on the fly.
